Design and Implementation of Probabilistic Programming Language Anglican

### 2016-11-20 paper/paper.tex
added self-citation to Output-Sensitive
describing the addressing scheme

\hy{This subsection is also not quite self-contained. Someone
who is not familiar with Clojure or Anglican codebase is likely
to get lost. On the other hand, it is a good overview or guide
for those who want to study the implemenation of Anglican. One
way to address this presentation issue while keeping the benefits
of the current text is to include some extended examples. For instance,
in the beginning of the paper, we might include several
Anglican examples. Then, we show that how these examples get
cps-transformed. Next, we present the implementation of inference
algorithms in Anglican using semi pseudo code; this is not quite
an implementation in Clojure but gives a good sketch on how
various functions are implemented in each inference algorithm.
Finally, we show how these algorithms work on cps-tranformed 
examples.}

>>>>

\hy{This subsection is also not quite self-contained. Someone
who is not familiar with Clojure or Anglican codebase is likely
to get lost. On the other hand, it is a good overview or guide
for those who want to study the implementation of Anglican. One
way to address this presentation issue while keeping the benefits
of the current text is to include some extended examples. For instance,
in the beginning of the paper, we might include several
Anglican examples. Then, we show that how these examples get
cps-transformed. Next, we present the implementation of inference
algorithms in Anglican using semi pseudo code; this is not quite
an implementation in Clojure but gives a good sketch on how
various functions are implemented in each inference algorithm.
Finally, we show how these algorithms work on cps-transformed 
examples.}

\subsection{Addressing of checkpoints}

Many inference algorithms memoize and reuse earlier computations
at checkpoints: variants of Metropolis-Hastings reuse previously
drawn values~\cite{WSG11}, as well as additional information
used for adaptive sampling~\cite{TMP+15} at \texttt{sample}
checkpoints, Asynchronous SMC (Particle Cascade) computes
average particle weights at \texttt{observe}
checkpoints~\cite{PWD+14}. Implementations of black-box
variational inference~\cite{WW13,MPT+16} associate with random
variables the learned parameters of variational posterior
distribution.

Checkpoints can be uniquely named at the compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple time due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially, however it impacts the 
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item The address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments 
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))
      x (sample dist)]
  (predict x))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces, however the random values should not be
re-used between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of 
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

To overcome the above problems, Anglican uses a different scheme
which is almost as efficient as the scheme based on stack addresses for
re-use of previously drawn values, produces addresses of constant size,
and allows manual computation of checkpoint addresses at runtime
when the default automatic scheme is insufficient.

\begin{verbatim}
A simple heuristic gives the same acceptance rate as with interpreted Anglican on all examples,
including Neil’s Locomotion, as follows:

  a) each sample form has a unique id, generated at compile time. It does anyway, and this 
      has almost zero cost.
  b) address is a pair 
        [id-of-sample-form number-of-previous-occurences]
      the occurences are of the same form.
  c) if a sequence of random choices with the same id is interrupted, the count is padded up 
  by a small power of 2, for example 16.

For example, if random choices

(sample S1)
(sample ‘S2 …
(sample ‘S2 …
(sample ‘S1 …
(sample ‘S1 …
(sample ‘S1 …
(sample ‘S2 …

are encountered in the trace, then the addresses are:

[S1 1]
[S2 1]
[S2 2]
[S1 16]
[S1 17]
[S1 18]
[S2 16]

This way, every occurence of random choice has a unique address, but small disturbances (like
removal or addition of a single choice of a particular kind) do not derail the entire sequence.
This aligns well with recursive patterns used in DPmem, for example.

I have some thoughts about stopping early in addition to starting late, but will do it when there
are less deadlines. Right now, the code runs MUCH MUCH faster, it should take roughly 10
hours to produce the same amount of data for Neil’s Locomotion example which was produced
during 14 DAYS of the holidays. 
\end{verbatim}
